---
title: "week1_exercise2_partb"
author: "Chen Chen"
format: html
embed-resources: true
editor: visual
---

## Week 1 Exercise Part B

### 1. Generating sampled values

We sample 100 values from a log-normal distribution (mu = 5, sigma = 0.5) using $rlnorm$ function. We also set seed to make it reproducible.

```{r}
set.seed(18)
samplevalue <- rlnorm(100, meanlog = 5, sdlog = 0.5)
```

### 2. Plotting

We create plots using $ggplot2$ package.

### 2.1 Histogram of the sampled data on linear scales

We first upload $ggplot2$ package, and create histogram on linear scales.

```{r}
library('ggplot2')
histolinear <- ggplot(data.frame(samplevalue), aes(x = samplevalue)) +
  geom_histogram(bins = 30, color = 'black', fill = 'lightblue') +
  labs(title = 'Histogram of the sampled data', x = 'Sample value', y = 'Count')

histolinear
```

### 2.2 Histogram of the sampled data on log scales

We create histogram on transformed sampled data (log scale).

```{r}
histolog <- ggplot(data.frame(samplevalue), aes(x = log(samplevalue))) +
  geom_histogram(bins = 30, color = 'black', fill = 'lightblue') +
  labs(title = 'Histogram of the sampled data (log scale)', x = 'Sample value (log scale)', y = 'Count')

histolog
```

### 2.3 Violin plot

We create a violin plot.

```{r}
violin<- ggplot(data.frame(samplevalue), aes(x = '', y = samplevalue)) +
  geom_violin() +
  labs(title = 'violin plot of the sampled data', x = 'Sample', y = 'Sample value')

violin
```

### 3. Reporting the mean and standard deviation of the sampled values

The mean of our sampled values is `r mean(samplevalue)` and the standard deviation of our sampled values is `r sd(samplevalue)` .
